Bayesian neural networks for sparse coding

### 2018-10-23 icassp_2019/paper.tex
attempt to shorten, forward propagation mainly successful

\begin{abstract}
  While deep learning has an overwhelming success, uncertainty estimation is still a challenge for most modern deep learning methods. The Bayesian learning allows to estimate the uncertainty of predictions in neural networks by imposing the prior distributions on weights, propagating the resulting uncertainty through the layers and computing the posterior distributions of predictions. We propose a novel method of propagating the uncertainty through the soft thresholding nonlinearity. The method expands the area of Bayesian neural networks and enables the soft thresholding nonlinearity to be included in Bayesian deep neural networks for the first time. As an example, we design a Bayesian Learned Iterative Shrinkage-Thresholding network (Bayesian \textsc{lista}). We develop a highly efficient posterior inference algorithm based on probabilistic backpropagation. A series of experiments on sparse coding shows that the proposed framework provides both accurate predictions and sensible estimates of uncertainty in these predictions.
\end{abstract}

@@

  In general, when distributions are included in the network, the complexity of Bayesian inference scales exponentially with the number of layers. Nevertheless, recently several techniques have been proposed to handle specific types of neural networks without this exponential growth. For example, feed-forward networks with the rectified linear unit nonlinearity~\cite{hernandez2015probabilistic}, networks with discrete distributions~\cite{soudry2014expectation}, recurrent networks~\cite{mcdermott2017bayesian}.

  There are currently two approaches in Bayesian neural networks: explicit distributions of weights can be included in a network \cite{hernandez2015probabilistic, ranganath2015deep}, or dropout can be interpreted as an element introducing uncertainty \cite{gal2016dropout}. We use the former approach and impose prior distributions on the neural network's weights to obtain the posterior distributions of weights and posterior predictive distributions of target variables.

@@

  In this paper, we propose a method to propagate uncertainty through the soft thresholding nonlinearity. At every layer the current distribution of the target vector is represented as a spike and slab distribution \cite{mitchell1988bayesian}, which represents the probabilities of each variable being zero, or Gaussian-distributed. We show that the spike and slab distribution of target vectors can be effectively combined with Gaussian weights of dense layers. The distribution of the target vector propagated through a dense layer and the soft thresholding nonlinearity can then be closely approximated with a distribution from the same family. The proposed method of uncertainty propagation for the soft thresholding nonlinearity allows us to derive the gradients of the logarithms of normalisation constants, which can be used to update a weight distribution. We then develop a highly efficient Bayesian inference algorithm based on the probabilistic backpropagation paradigm while significantly expanding it.

  The main contributions of this paper are: (\textit{i}) for the first time we propose a method for uncertainty propagation through the soft thresholding nonlinearity for a Bayesian neural network; (\textit{ii}) we develop an efficient posterior inference algorithm for weights and outputs of neural networks with the soft thresholding nonlinearity; (\textit{iii}) we design a novel Bayesian neural network for sparse coding.

  We introduce a novel Bayesian neural network in section \ref{sec:bayesian_lista} after reviewing neural networks for sparse coding in section \ref{sec:nn_sc}. The proposed forward uncertainty propagation and probabilistic backpropagation methods are given in sections \ref{sec:fprop} and \ref{sec:backpropagation}. We show the experimental results in section \ref{sec:experiments} and finish with discussion and conclusions in section \ref{sec:conclusions}.

\section{Neural networks for sparse coding}
\label{sec:nn_sc}
  The neural network approach to sparse coding is based on earlier Iterative Shrinkage and Thresholding Algorithm (\textsc{ista}) \cite{daubechies2004iterative}. It addresses the sparse coding problem as the linear regression problem with the $l1$ penalty that promotes sparsity. For the linear regression model with observations $\mathbf{y} \in \mathbb{R}^K$, the design matrix $\mathbf{X} \in \mathbb{R}^{K \times D}$, and the sparse unknown vector of weights $\boldsymbol\beta \in \mathbb{R}^D$, \textsc{ista} minimises the energy function $E(\cdot)$ to find $\boldsymbol\beta$
  \begin{equation}
  \label{eq:regression_problem}
  E(\mathbf{y}, \boldsymbol\beta) = ||\mathbf{X}\boldsymbol\beta - \mathbf{y}||_2^2 + \alpha ||\boldsymbol\beta||_1,
  \end{equation}
  where $\alpha$ is the parameter that controls the influence of the sparse penalty.

  At every iteration $l$, \textsc{ista} obtains the new estimate $\widehat{\boldsymbol\beta}_l$ of the target vector $\boldsymbol\beta$ as the linear transformation $\mathbf{b} = \mathbf{W}\mathbf{y} + \mathbf{S}\widehat{\boldsymbol\beta}_{l-1}$ that is propagated through the soft thresholding function $h_\lambda(\cdot)$
  \begin{equation}
  h_\lambda(\mathbf{b}) = \text{sgn}(\mathbf{b}) \max(|\mathbf{b}| - \lambda, 0),
  \end{equation}
  where $\lambda$ is a shrinkage parameter.
  In \textsc{ista}, weights of the linear transformation are assumed fixed: $\mathbf{W} = \mathbf{X}^\top / E$, where $E$ is the upper bound of the largest eigenvalue of $\mathbf{X}^\top\mathbf{X}$, and $\mathbf{S} = \mathbf{I}_{D \times D} - \mathbf{W}\mathbf{X}$, where $\mathbf{I}_{D \times D}$ is the identity matrix of size $D$.

  In contrast to \textsc{ista}, Learned \textsc{ista} (\textsc{lista}) \cite{gregor2010learning} learns the values of matrices $\mathbf{W}$, $\mathbf{S}$ based on a set of pairs $\{\mathbf{Y}, \mathbf{B}\}=\{\mathbf{y}^{(n)}, \boldsymbol\beta^{(n)}\}_{n=1}^N$, where $N$ is the number of these pairs. To achieve this, the \textsc{ista} is limited with the fixed amount of iterations $L$ and interpreted as a recurrent neural network: every iteration $l$ of \textsc{ista} corresponds to the layer $l$ of \textsc{lista}. A coefficient vector $\boldsymbol\beta$ for an observation~$\mathbf{y}$ is predicted according to Algorithm~\ref{alg:lista}.

  \begin{algorithm}[h]
    \caption{\textsc{lista} forward propagation}
    \label{alg:lista}
    \begin{algorithmic}[1]
      \REQUIRE observations $\mathbf{y}$, weights $\mathbf{W}, \mathbf{S}$, number of layers $L$
      \STATE \textit{Initialisation.} Dense layer $\mathbf{b} \gets \mathbf{W}\mathbf{y}$ \label{eq:first_layer}
      \STATE \textit{Initialisation.} Soft thresholding nonlinearity $\widehat{\boldsymbol\beta}_0 \gets h_\lambda(\mathbf{b})$ \label{eq:thr_first}
      \FOR{$l=1$ \TO $L$}
        \STATE Dense layer $\mathbf{c}_l \gets \mathbf{b} + \mathbf{S}\widehat{\boldsymbol\beta}_{l-1}$ \label{eq:l_dense_layer}
        \STATE Soft thresholding nonlinearity $\widehat{\boldsymbol\beta}_{l} \gets h_\lambda(\mathbf{c}_l)$ \label{eq:l_thr}
      \ENDFOR
      \RETURN $\widehat{\boldsymbol\beta} \gets \widehat{\boldsymbol\beta}_{L}$
    \end{algorithmic}
  \end{algorithm}

  In \textsc{lista}, matrices $\mathbf{W}$, $\mathbf{S}$ are the parameters that are initialised as in the \textsc{ista} and then updated with the backpropagation algorithm. Vectors $\mathbf{c}_l$ and $\mathbf{b}$ are auxiliary vectors that describe forward propagation.

  \section{Bayesian neural network for sparse coding}
  \label{sec:bayesian_lista}
  To formulate the Bayesian version of \textsc{lista} we impose the prior distributions on the unknown weights
  \begin{equation}
  \label{eq:ws}
  p(\mathbf{W}) = \prod_{d=1}^D\prod_{k=1}^K \mathcal{N}(w_{ij} ; 0, \eta^{-1}), \qquad
  p(\mathbf{S}) = \prod_{d'=1}^D\prod_{d''=1}^D \mathcal{N}(s_{d'd''} ; 0, \eta^{-1}),
  \end{equation}
  where $\eta$ is the precision of the Gaussian distribution.

  For every layer $l$ of Bayesian \textsc{lista}, we assume that $\widehat{\boldsymbol\beta}_{l}$ has the spike and slab distribution with the spike probability $\boldsymbol\omega$, the slab mean $\mathbf{m}$, and the slab variance $\mathbf{v}$
  \begin{equation}
  [\widehat{\boldsymbol\beta}_{l}]_d \sim \omega_d \delta_0 + (1 - \omega_d)\mathcal{N}(m_d, v_d),
  \end{equation}
  where $\delta_0$ is the delta-function that represents a spike, $[\cdot]_d$ denotes the $d$-th component of a vector. In section \ref{sec:fprop}, we show that the output of the next layer $\widehat{\boldsymbol\beta}_{l+1}$ can be approximated with a spike and slab distribution and, therefore, the output of the Bayesian \textsc{lista} network $\widehat{\boldsymbol\beta}$ has the spike and slab distribution.

  To introduce the uncertainty of predictions, we assume that the true $\boldsymbol\beta$ is an output $f(\mathbf{y} ; \mathbf{S}, \mathbf{W}, \lambda)$ of the Bayesian \textsc{lista} network corrupted by the additive Gaussian zero-mean noise with the precision~$\gamma$. Then the likelihood of $\mathbf{B}$ is defined as
  \begin{equation}
  \label{eq:likelihood}
  p(\mathbf{B}| \mathbf{Y}, \mathbf{W}, \mathbf{S}, \gamma, \lambda) = \prod_{n=1}^N\prod_{d=1}^D\mathcal{N}\left(\beta_d^{(n)}; [f(\mathbf{y} ; \mathbf{S}, \mathbf{W}, \lambda)]_d, \gamma^{-1}\right)
  \end{equation}
  We specify Gamma prior distributions with parameters~$a^{\cdot}$ and~$b^{\cdot}$ on the introduced Gaussian precisions
  \begin{equation}
  \label{eq:gamma_eta}
  p(\gamma) = \text{Gam}\left(\gamma; a^{\gamma}, b^{\gamma}\right), \qquad
  p(\eta) = \text{Gam}\left(\eta; 	a^{\eta}, b^{\eta}\right)
  \end{equation}

  The posterior distribution is then
  \begin{equation}
  \label{eq:posterior}
  p(\mathbf{W}, \mathbf{S}, \gamma, \eta | \mathbf{B}, \mathbf{Y}, \lambda) = \frac{p(\mathbf{B} | \mathbf{Y}, \mathbf{W},  \mathbf{S}, \gamma, \lambda) p(\mathbf{W} | \eta )p(\mathbf{S} | \eta) p(\eta) p(\gamma)}{p(\mathbf{B} | \mathbf{Y}, \lambda)}
  \end{equation}
  The shrinkage parameter $\lambda$ is a hyperparameter of the model.

  \section{Uncertainty propagation through soft thresholding}
  \label{sec:fprop}
  We modify \textsc{lista} forward propagation (Algorithm~\ref{alg:lista}) to include probability distributions of the random variables introduced in section~\ref{sec:bayesian_lista}.
  We demonstrate that the output of every Bayesian \textsc{lista} layer $\widehat{\boldsymbol\beta}_l$ can be closely approximated with a spike and slab distribution. This leads to the proposed probabilistic backpropagation algorithm that is presented in section~\ref{sec:backpropagation}.

  \subsection{Initialisation}
  At step \ref{eq:first_layer} of \textsc{lista} (Algorithm~\ref{alg:lista}) the matrix $\mathbf{W}$ consists of Gaussian-distributed components $w_{dk} \sim \mathcal{N}(m^w_{dk}, v^w_{dk})$, and $\mathbf{y}$ is a deterministic vector. Since the sum of independent Gaussian-distributed variables is a Gaussian variable~\cite{eisenberg2008sum}, $\mathbf{b}$ is a vector of Gaussian-distributed components $b_d \sim \mathcal{N}(m^b_d, v^b_d)$
  \begin{equation}
  \label{eq:matrix_vector_product}
  m^b_d = \sum_{k=1}^Ky_k m^w_{dk}, \qquad
  v^b_d = \sum_{k=1}^Ky_k^2v^w_{dk}.
  \end{equation}

  At step~\ref{eq:thr_first} of \textsc{lista} (Algorithm~\ref{alg:lista}) the Gaussian vector $\mathbf{b}$ is taken as an input of the soft thresholding function.

  When a Gaussian random variable $x \sim \mathcal{N}(x; m, v)$ is propagated through the soft thresholding function $x^* = h_{\lambda}(x)$, the probability mass of the resulting random variable $x^*$ is split into two parts. The values of $x$ from the interval $[-\lambda, \lambda]$ are converted to~$0$ by the soft thresholding operator. Therefore, the probability mass of the original distribution that lies in $[-\lambda, \lambda]$ is squeezed into the probability of $x^*$ being zero. The values of $x$ from outside of the $[-\lambda, \lambda]$ interval are shifted towards~$0$. The distribution of $x^* \neq 0$ then represents the tails of the original Gaussian distribution.

  \begin{lemma}[Propagation of a Gaussian variable through soft thresholding]
  \label{thm:soft_thresholding}
  The distribution of $x^*$ can be parametrised by the probability of being zero, $\omega^*$, the mean $m^*$ and the variance $v^*$ of the truncated Gaussian distribution.
  \end{lemma}

  The proof of Lemma~\ref{thm:soft_thresholding} and expressions for $\omega^*$, $m^*$ and $v^*$ are given in the supplementary materials. Based on the results of Lemma~\ref{thm:soft_thresholding}, we approximate the distribution of $\widehat{\boldsymbol\beta}_0$ at step~\ref{eq:thr_first} with a spike and slab distribution with parameters derived in the lemma: the spike probability $\omega$ is equal to $\omega^*$, the slab mean $m$ and variance $v$ are set to the corresponding parameters of the truncated Gaussian $m^*$ and $v^*$. The quality of this approximation is discussed in the supplementary materials.

  \subsection{Main layers}
  At step \ref{eq:l_dense_layer} of \textsc{lista} (Algorithm~\ref{alg:lista}) the vector $\mathbf{b}$ and matrix $\mathbf{S}$ consist of Gaussian components and $\widehat{\boldsymbol\beta}_{l-1}$ is a vector of the spike and slab random variables. In order to determine the distribution of the output $\mathbf{c}_l$ we first formulate two lemmas.

  \begin{lemma}[Moments of a spike and slab distribution]
  \label{thm:moments_spsl}
  Let a random variable $\xi$ have a spike and slab distribution with the probability of spike $\omega$, the slab mean $m$ and slab variance $v$. Then its moments are
  \begin{equation}
  \mathbb{E}\xi = (1-\omega)m, \qquad \operatorname{Var}\xi = (1-\omega)(v + \omega m^2)
  \end{equation}
  \end{lemma}

  \begin{lemma}[Product of a Gaussian matrix and a spike and slab vector]
    \label{thm:matrix_vector}
  Let $\mathbf{S} \in \mathbb{R}^{D \times D}$ be a matrix of independent Gaussian-distributed random variables: $s_{d'd''} \sim \mathcal{N}(m^s_{d'd''}, v^s_{d'd''})$, and $\widehat{\boldsymbol\beta }\in \mathbb{R}^D$ be a vector with spike-and-slab-distributed variables: $\widehat{\beta}_d \sim \omega_d \delta_0 + (1 - \omega_d)\mathcal{N}(m_d, v_d)$. The components of the matrix $\mathbf{S}$ and the vector $\widehat{\boldsymbol\beta}$ are mutually independent. Let $\mathbf{e} \in \mathbb{R}^{D}$ denote the product $\mathbf{S} \widehat{\boldsymbol\beta}$. Then the marginal mean and variance of the elements $e_d$ of the vector $\mathbf{e}$ are
  \begin{subequations}
  \begin{align}
   \mathbb{E}e_d &= \sum_{d'=1}^D m^s_{dd'}(1-\omega_{d'})m_{d'}, \\
   \begin{split}
   \operatorname{Var}e_d &= \sum_{d'=1}^D [(m^s_{dd'})^2(1-\omega_{d'})^2v_{d'} + (1-\omega_{d'})^2(m_{d'})^2v^s_{dd'} + v^s_{dd'}(1-\omega_{d'})^2v_{d'}].
   \end{split}
   \end{align}
  \end{subequations}
   \end{lemma}

  The proofs of Lemmas~\ref{thm:moments_spsl} and~\ref{thm:matrix_vector} are given in the supplementary materials.

  Let $\mathbf{e}_l$ denote the product $\mathbf{S}\widehat{\boldsymbol\beta}_{l-1}$ at step \ref{eq:l_dense_layer} of \textsc{lista}. We assume that $\mathbf{S}$ and $\widehat{\boldsymbol\beta}_{l-1}$ are mutually independent. Then according to the Central Limit Theorem $[ \mathbf{e}_l ]_d$ can be approximated as a Gaussian-distributed variable when $D$ is sufficiently large. The parameters of the of the approximating Gaussian distribution are set to the marginal mean and variance given in Lemma~\ref{thm:matrix_vector}. The supplementary materials provide a discussion of the quality of this approximation.

  The output $\mathbf{c}_l$ at step \ref{eq:l_dense_layer} is then represented as a sum of two Gaussian-distributed vectors: $\mathbf{b}$ and~$\mathbf{e}_l$, i.e. it is a Gaussian-distributed vector with components $c_{d} \sim \mathcal{N}(m^c_{d}, v^c_{d})$
  \begin{equation}
  \label{eq:sum_vectors}
  m^c_{d} = m^b_{d} + m^e_{d}, \qquad
  v^c_{d} = v^b_{d} + v^e_{d}.
  \end{equation}

  Then $\widehat{\boldsymbol\beta}_{l}$ at step \ref{eq:l_thr} of \textsc{lista} (Algorithm~\ref{alg:lista}) is the result of soft thresholding of a Gaussian variable, which is approximated with the spike and slab distribution,  similar to step \ref{eq:thr_first}.

  \subsection{Bayesian \textsc{lista} forward propagation}
  We have covered all steps of Bayesian \textsc{lista} and derived distributions for outputs of these steps. Now we can formulate the main result of this section that describes how the proposed uncertainty propagation scheme works.

>>>>

\begin{abstract}
  While deep learning has an overwhelming success, uncertainty estimation is still a challenge for most modern deep learning methods. Bayesian learning allows to estimate the uncertainty of predictions in neural networks by imposing the prior distributions on weights, propagating the resulting uncertainty through the layers and computing the posterior distributions of predictions. We propose a novel method of propagating the uncertainty through the soft thresholding nonlinearity. The method expands the area of Bayesian neural networks and enables the soft thresholding nonlinearity to be included in Bayesian deep neural networks for the first time. As an example, we design a Bayesian Learned Iterative Shrinkage-Thresholding network (Bayesian \textsc{lista}). An efficient posterior inference algorithm based on probabilistic backpropagation is developed. A series of experiments on sparse coding shows that the proposed framework provides both accurate predictions and sensible estimates of uncertainty in these predictions.
\end{abstract}

@@

  In general, when distributions are included in a network, the complexity of Bayesian inference scales exponentially with the number of layers. Nevertheless, recently several techniques~\cite{ranganath2015deep, gal2016dropout} have been proposed to handle specific types of neural networks without this exponential growth. For example, feed-forward networks with the rectified linear unit nonlinearity~\cite{hernandez2015probabilistic}, networks with discrete distributions~\cite{soudry2014expectation}, recurrent networks~\cite{mcdermott2017bayesian}.

%  There are currently two approaches in Bayesian neural networks: explicit distributions of weights can be included in a network \cite{hernandez2015probabilistic, ranganath2015deep}, or dropout can be interpreted as an element introducing uncertainty \cite{gal2016dropout}. We use the former approach and impose prior distributions on the neural network's weights to obtain the posterior distributions of weights and posterior predictive distributions of target variables.

@@

A novel method to propagate uncertainty through the soft thresholding nonlinearity is proposed in this paper. At every layer the current distribution of the target vector is represented as a spike and slab distribution \cite{mitchell1988bayesian}, which represents the probabilities of each variable being zero, or Gaussian-distributed. 
%We show that the spike and slab distribution of target vectors can be effectively combined with Gaussian weights of dense layers. The distribution of the target vector propagated through a dense layer and the soft thresholding nonlinearity can then be closely approximated with a distribution from the same family. 
Using the proposed method of uncertainty propagation for the soft thresholding nonlinearity, the gradients of the logarithms of normalisation constants are derived, which can be used to update a weight distribution. An efficient Bayesian inference algorithm is then developed based on the probabilistic backpropagation paradigm while significantly expanding it. A novel Bayesian neural network for sparse coding is designed utilising both the proposed method of uncertainty propagation and Bayesian inference algorithm.

  The main contributions of this paper are: (\textit{i}) for the first time a method for uncertainty propagation through the soft thresholding nonlinearity is proposed for a Bayesian neural network; (\textit{ii}) an efficient posterior inference algorithm for weights and outputs of neural networks with the soft thresholding nonlinearity is developed; (\textit{iii}) a novel Bayesian neural network for sparse coding is designed.

 The rest of the paper is organised as follows.  A novel Bayesian neural network is introduced in section \ref{sec:bayesian_lista}. The proposed forward uncertainty propagation and probabilistic backpropagation methods are given in sections \ref{sec:fprop} and \ref{sec:backpropagation}. Section \ref{sec:experiments} provides the experimental results and section \ref{sec:conclusions} concludes the paper.

\section{Bayesian neural network for sparse coding}
  \label{sec:bayesian_lista}
  This section first reviews existing neural networks for sparse coding and then describes the novel Bayesian neural network.

\subsection{Frequentist neural networks}  
\label{subsec:nn_sc}
  The neural network approach to sparse coding is based on earlier Iterative Shrinkage and Thresholding Algorithm (\textsc{ista}) \cite{daubechies2004iterative}. It addresses the sparse coding problem as the linear regression problem with the $l1$ penalty that promotes sparsity. For the linear regression model with observations $\mathbf{y} \in \mathbb{R}^K$, the design matrix $\mathbf{X} \in \mathbb{R}^{K \times D}$, and the sparse unknown vector of weights $\boldsymbol\beta \in \mathbb{R}^D$, \textsc{ista} minimises the energy function $E(\cdot)$ to find $\boldsymbol\beta$
  \begin{equation}
  \label{eq:regression_problem}
  E(\mathbf{y}, \boldsymbol\beta) = ||\mathbf{X}\boldsymbol\beta - \mathbf{y}||_2^2 + \alpha ||\boldsymbol\beta||_1,
  \end{equation}
  where $\alpha$ is the parameter that controls the influence of the sparse penalty.

  At every iteration $l$, \textsc{ista} obtains the new estimate $\widehat{\boldsymbol\beta}_l$ of the target vector $\boldsymbol\beta$ as the linear transformation $\mathbf{b} = \mathbf{W}\mathbf{y} + \mathbf{S}\widehat{\boldsymbol\beta}_{l-1}$ that is propagated through the soft thresholding function $h_\lambda(\cdot)$
  \begin{equation}
  h_\lambda(\mathbf{b}) = \text{sgn}(\mathbf{b}) \max(|\mathbf{b}| - \lambda, 0),
  \end{equation}
  where $\lambda$ is a shrinkage parameter.
  In \textsc{ista}, weights of the linear transformation are assumed fixed: $\mathbf{W} = \mathbf{X}^\top / E$, where $E$ is the upper bound of the largest eigenvalue of $\mathbf{X}^\top\mathbf{X}$, and $\mathbf{S} = \mathbf{I}_{D \times D} - \mathbf{W}\mathbf{X}$, where $\mathbf{I}_{D \times D}$ is the identity matrix of size~$D$.

  In contrast to \textsc{ista}, Learned \textsc{ista} (\textsc{lista}) \cite{gregor2010learning} learns the values of matrices $\mathbf{W}$ and $\mathbf{S}$ based on a set of pairs $\{\mathbf{Y}, \mathbf{B}\}=\{\mathbf{y}^{(n)}, \boldsymbol\beta^{(n)}\}_{n=1}^N$, where $N$ is the number of these pairs. To achieve this, the \textsc{ista} is limited with the fixed amount of iterations $L$ and interpreted as a recurrent neural network: every iteration $l$ of \textsc{ista} corresponds to the layer $l$ of \textsc{lista}. A coefficient vector $\boldsymbol\beta$ for an observation~$\mathbf{y}$ is predicted according to Algorithm~\ref{alg:lista}.

  \begin{algorithm}[t]
    \caption{\textsc{lista} forward propagation}
    \label{alg:lista}
    \begin{algorithmic}[1]
      \REQUIRE observations $\mathbf{y}$, weights $\mathbf{W}, \mathbf{S}$, number of layers $L$
      \STATE Dense layer $\mathbf{b} \gets \mathbf{W}\mathbf{y}$ \label{eq:first_layer}
      \STATE Soft thresholding nonlinearity~$\widehat{\boldsymbol\beta}_0 \gets h_\lambda(\mathbf{b})$ \label{eq:thr_first}
      \FOR{$l=1$ \TO $L$}
        \STATE Dense layer $\mathbf{c}_l \gets \mathbf{b} + \mathbf{S}\widehat{\boldsymbol\beta}_{l-1}$ \label{eq:l_dense_layer}
        \STATE Soft thresholding nonlinearity $\widehat{\boldsymbol\beta}_{l} \gets h_\lambda(\mathbf{c}_l)$ \label{eq:l_thr}
      \ENDFOR
      \RETURN $\widehat{\boldsymbol\beta} \gets \widehat{\boldsymbol\beta}_{L}$
    \end{algorithmic}
  \end{algorithm}

  In \textsc{lista}, matrices $\mathbf{W}$ and $\mathbf{S}$ are the parameters that are initialised as in the \textsc{ista} and then updated with the backpropagation algorithm. Vectors $\mathbf{c}_l$ and $\mathbf{b}$ are auxiliary vectors that describe forward propagation.

  \subsection{Baysian \textsc{lista}}
  \label{subsec:bayesian_lista}
  This section introduces the proposed Bayesian version of \textsc{lista}. The prior distributions are imposed on the unknown weights
  \begin{equation}
  \label{eq:ws}
  \begin{split}
  p(\mathbf{W}) &= \prod_{d=1}^D\prod_{k=1}^K \mathcal{N}(w_{ij} ; 0, \eta^{-1}), \\
  p(\mathbf{S}) &= \prod_{d'=1}^D\prod_{d''=1}^D \mathcal{N}(s_{d'd''} ; 0, \eta^{-1}),
  \end{split}
  \end{equation}
  where $\eta$ is the precision of the Gaussian distribution.

  For every layer $l$ of Bayesian \textsc{lista}, $\widehat{\boldsymbol\beta}_{l}$ is assumed to have the spike and slab distribution with the spike probability $\boldsymbol\omega$, the slab mean $\mathbf{m}$, and the slab variance $\mathbf{v}$
  \begin{equation}
  [\widehat{\boldsymbol\beta}_{l}]_d \sim \omega_d \delta_0 + (1 - \omega_d)\mathcal{N}(m_d, v_d),
  \end{equation}
  where $\delta_0$ is the delta-function that represents a spike, $[\cdot]_d$ denotes the $d$-th component of a vector. Section \ref{sec:fprop} shows that the output of the next layer $\widehat{\boldsymbol\beta}_{l+1}$ can be approximated with a spike and slab distribution and, therefore, the output of the Bayesian \textsc{lista} network $\widehat{\boldsymbol\beta}$ has the spike and slab distribution.

  To introduce the uncertainty of predictions, we assume that the true $\boldsymbol\beta$ is an output $f(\mathbf{y} ; \mathbf{S}, \mathbf{W}, \lambda)$ of the Bayesian \textsc{lista} network corrupted by the additive Gaussian zero-mean noise with the precision~$\gamma$. Then the likelihood of $\mathbf{B}$ is defined as
  \begin{equation}
  \label{eq:likelihood}
  \begin{split}
  p(\mathbf{B}| \mathbf{Y}, &\mathbf{W}, \mathbf{S}, \gamma, \lambda) \\
  = &\prod_{n=1}^N\prod_{d=1}^D\mathcal{N}\left(\beta_d^{(n)}; [f(\mathbf{y} ; \mathbf{S}, \mathbf{W}, \lambda)]_d, \gamma^{-1}\right)
 \end{split}
  \end{equation}
  Gamma prior distributions with parameters~$a^{\cdot}$ and~$b^{\cdot}$ are specified on the introduced Gaussian precisions
  \begin{equation}
  \label{eq:gamma_eta}
  p(\gamma) = \text{Gam}\left(\gamma; a^{\gamma}, b^{\gamma}\right), \qquad
  p(\eta) = \text{Gam}\left(\eta; 	a^{\eta}, b^{\eta}\right)
  \end{equation}

  The posterior distribution is then
  \begin{equation}
  \label{eq:posterior}
  \begin{split}
  p(\mathbf{W}, \mathbf{S}, &\gamma, \eta | \mathbf{B}, \mathbf{Y}, \lambda) \\
  = &\frac{p(\mathbf{B} | \mathbf{Y}, \mathbf{W},  \mathbf{S}, \gamma, \lambda) p(\mathbf{W} | \eta )p(\mathbf{S} | \eta) p(\eta) p(\gamma)}{p(\mathbf{B} | \mathbf{Y}, \lambda)}
  \end{split}
  \end{equation}
  The shrinkage parameter $\lambda$ is a hyperparameter of the model.

  \section{Uncertainty propagation through soft thresholding}
  \label{sec:fprop}
  This section describes modification of \textsc{lista} forward propagation (Algorithm~\ref{alg:lista}) to include probability distributions of the random variables introduced in section~\ref{sec:bayesian_lista}.

  \subsection{Initialisation}
  \label{subsec:forward_init}
  At step \ref{eq:first_layer} of \textsc{lista} (Algorithm~\ref{alg:lista}) the matrix $\mathbf{W}$ consists of Gaussian-distributed components $w_{dk} \sim \mathcal{N}(m^w_{dk}, v^w_{dk})$, and $\mathbf{y}$ is a deterministic vector. Then the output $\mathbf{b}$ is a vector of Gaussian-distributed components $b_d \sim \mathcal{N}(m^b_d, v^b_d)$, where $m^b_d = \sum_{k=1}^Ky_k m^w_{dk}$, and $v^b_d = \sum_{k=1}^Ky_k^2v^w_{dk}$.
%  \begin{equation}
%  \label{eq:matrix_vector_product}
%  m^b_d = \sum_{k=1}^Ky_k m^w_{dk}, \qquad
%  v^b_d = \sum_{k=1}^Ky_k^2v^w_{dk}.
%  \end{equation}

  At step~\ref{eq:thr_first} of \textsc{lista} (Algorithm~\ref{alg:lista}) the Gaussian vector $\mathbf{b}$ is taken as an input of the soft thresholding function. When a Gaussian random variable $x \sim \mathcal{N}(x; m, v)$ is propagated through the soft thresholding function $x^* = h_{\lambda}(x)$, the probability mass of the resulting random variable $x^*$ is split into two parts. The values of $x$ from the interval $[-\lambda, \lambda]$ are converted to~$0$ by the soft thresholding operator. Therefore, the probability mass of the original distribution that lies in $[-\lambda, \lambda]$ is squeezed into the probability of $x^*$ being zero. The values of $x$ from outside of the $[-\lambda, \lambda]$ interval are shifted towards~$0$. The distribution of $x^* \neq 0$ then represents the tails of the original Gaussian distribution. The distribution of $x^*$ can be then parametrised by the probability of being zero, $\omega^*$, the mean $m^*$ and the variance $v^*$ of the truncated Gaussian distribution.
%  \begin{lemma}[Propagation of a Gaussian variable through soft thresholding]
%  \label{thm:soft_thresholding}
%  The distribution of $x^*$ can be parametrised by the probability of being zero, $\omega^*$, the mean $m^*$ and the variance $v^*$ of the truncated Gaussian distribution.
%  \end{lemma}
 %Based on the results of Lemma~\ref{thm:soft_thresholding}, 
 Therefore, we approximate the distribution of $\widehat{\boldsymbol\beta}_0$ at step~\ref{eq:thr_first} with a spike and slab distribution with the corresponding parameters: the spike probability $\omega^*$, the slab mean $m^*$ and variance $v^*$.

  \subsection{Main layers}
  At step \ref{eq:l_dense_layer} of \textsc{lista} (Algorithm~\ref{alg:lista}) the vector $\mathbf{b}$ and matrix $\mathbf{S}$ consist of Gaussian components: $b_d \sim \mathcal{N}(m^b_d, v^b_d)$, $s_{d'd''} \sim \mathcal{N}(m^s_{d'd''}, v^s_{d'd''})$, and $\widehat{\boldsymbol\beta}_{l-1}$ is a vector of the spike and slab random variables: $[\widehat{\beta}_{l-1}]_d \sim \omega_d \delta_0 + (1 - \omega_d)\mathcal{N}(m_d, v_d)$.

 It can be shown that the expected value and variance of a spike and slab distributed variable $\xi$ with the probability of spike $\omega$, the slab mean $m$ and slab variance $v$ are:  
   \begin{equation}
   \label{eq:spsl_moments}
  \mathbb{E}\xi = (1-\omega)m, \qquad \operatorname{Var}\xi = (1-\omega)(v + \omega m^2)
  \end{equation}

 It can also be shown that if components of the matrix $\mathbf{S}$ and vector $\widehat{\boldsymbol\beta}_{l-1}$ are mutually independent then the components $[ \mathbf{e}_l ]_d$ of their product $\mathbf{e}_l = \mathbf{S} \widehat{\boldsymbol\beta}_{l-1}$ have the marginal mean and variances:
  \begin{subequations}
  \label{eq:e_moments}
  \begin{align}
  m^e_{d} \stackrel{\text{def}}{=} &\mathbb{E}[ \mathbf{e}_l ]_d = \sum_{d'=1}^D m^s_{dd'}(1-\omega_{d'})m_{d'}, \\
   \begin{split}
  v^e_{d} \stackrel{\text{def}}{=} &\operatorname{Var}[ \mathbf{e}_l ]_d = \sum_{d'=1}^D [(m^s_{dd'})^2(1-\omega_{d'})^2v_{d'} \\
   + &(1-\omega_{d'})^2(m_{d'})^2v^s_{dd'} + v^s_{dd'}(1-\omega_{d'})^2v_{d'}].
   \end{split}
   \end{align}
  \end{subequations}

 If the matrix $\mathbf{S}$ and vector $\widehat{\boldsymbol\beta}_{l-1}$ are assumed to be mutually independent, then according to the Central Limit Theorem $[ \mathbf{e}_l ]_d$ can be approximated as a Gaussian-distributed variable when $D$ is sufficiently large. The parameters of the of the approximating Gaussian distribution are set to the marginal mean and variance given in~(\ref{eq:e_moments}).
